Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-07-10 Manuscript_Nater_etal.qmd
Smaller text edits

Quantifying temporal and spatial variation in animal population size and demography is a central theme in ecological research and important for directing management and policy. However, this requires field sampling at large spatial extents and over long periods of time, which is not only prohibitively costly but often politically untenable. Participatory monitoring programs (also called citizen science programmes) can alleviate these constraints by recruiting stakeholders and the public to increase the spatial and temporal resolution of sampling effort and hence resulting data. While the majority of participatory monitoring programs are limited by opportunistic sampling designs, we are starting to see the emergence of structures citizen science programs that employ trained volunteers to collect data according to standardized protocols. Simultaneously, there is much ongoing development of statistical models that are increasingly more powerful and able to make more efficient use of field data. Integrated population models (IPMs), for example, are able to use multiple streams of data from different field monitoring programmes and/or multiple aspects of single datasets to estimate population sizes and key vital rates. Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale participatory monitoring program -- the "Hønsefuglportalen" -- to study spatio-temporal variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway. We constructed an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence. Recruitment rates varied more across space than over time, while the opposite was the case for survival. Slower life history patterns (higher survival, lower recruitment) appeared to be more common at higher latitudes and altitudes, portending differential effects of climate change on ptarmigan across their range. While there was variation in the magnitude of the effect small rodent occupancy had on recruitment, the relationships were predominantly positive and thus consistent with the alternative prey hypothesis. Notably, the accurate estimation of covariate effect was only made possible by integrating data from several monitoring areas for analysis. Our study highlights the potential of participatory monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns in species abundance and demographic rates, and showcases how corresponding workflows can be set up in a reproducible and semi-automated way that increases their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.

@@

While large-scale, spatially-explicit indicators for species abundance and populations are clearly needed, development and practical implementation are greatly limited due to the reliance of such indicators on the availability of data from large-scale, long-term monitoring programmes [@proencca2017]. Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last decades. Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), the PanEuropean Common Bird Monitoring Scheme (<https://pecbms.info/>), the UK Butterfly Monitoring Scheme (<https://ukbms.org/>), the Game and Wildlife Conservation Trust Partridge Count Scheme [@aebischer2010], and the Swiss Biodiversity Monitoring (<https://www.biodiversitymonitoring.ch/>).

@@

The line transect data from "Hønsefuglportalen" has been used previously for estimating abundance trends of willow ptarmigan (*Lagopus lagopus*) across Norway [e.g. @bowler2020; @nilsen2020], and to test a range of relevant ecological hypotheses [@bowler2020; @Breisjøberget_etal_2018]. However, large-scale estimation of demographic rates underlying abundance trends has thus far remained as untapped potential of the dataset. @nilsen2023 recently developed a novel integrated distance sampling model (IDSM) which successfully uses the age of individuals detected along line transects data coupled with radio-telemetry data to jointly estimate abundance, survival, and recruitment across years. In this study, we adapt and extend the model of @nilsen2023 to run not just on a single site but on all areas with publicly available line transect data from "Hønsefuglportalen" simultaneously. Unlike several previous studies applying integrated models for population dynamics to multiple (sub-) populations separately and comparing results [e.g. @robinson2014; @nater2023], we opt for an approach explicitly integrating across space, thus allowing for sharing of information across locations and -- in effect -- space-for-time substitution [e.g. @horswill2019; @morrison2022]. We then apply the resulting multi-area IDSM to "Hønsefuglportalen" data on willow ptarmigan to estimate population size, age-structure, survival, recruitment, and impacts of small rodent occupancy across 41 reporting districts and 15 years (2007-2021) for this culturally important small-game species. We further embed the modelling workflow in a reproducible, semi-automated pipeline that will greatly facilitate the repeated calculation of abundance and population indicators at different spatial scales as new data are added every year.

@@

The willow ptarmigan is a tetraoid bird with a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g. @fuglei2020]. While the species is currently listed as Least Concern (LC) both in the global and Norwegian Red List of Species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022]. The main reason for the long-term decline in abundance remain unresolved, but the willow ptarmigan are considered sentinel species that are sensitive to both climate change and land use changes [@Henden2017; @Storch2007]. Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for resident predators, such as the gyrfalcon [@Alastair2020]. The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can display substantial spatio-temporal variation in demographic rates [@bowler2020]. Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @bowler2020]. This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis (APH) and has been a central part of research on Fennoscandian grouse population dynamics for many decades [@Elton1942; @hagen1952; @Linden1988; @Steen1988]. In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023; @Steen1988]. Across their distributional range, willow ptarmigan are an iconic species with a high cultural value, partly linked to their popularity as game species. The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies [@Eriksen2023]. In addition, being a sentinel species, the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway, @jakobsson2020, <https://www.naturindeks.no/Indicators/lirype>] and ecosystem condition [Assessment of the Ecological Condition, @framstad2022] assessments.

@@

The line transect survey data were collected through a structured participatory monitoring program called "Hønsefuglportalen" (<https://honsefugl.nina.no/Innsyn/en>). In the first three weeks of August each year, trained volunteer fieldworkers collect observations of willow ptarmigan and other grouse species (rock ptarmigan *Lagopus muta*, black grouse *Lyrurus tetrix*, and capercaillie *Tetrao urogallus*) along predefined line transects. To increase the detection probability, fieldworkers use pointing dogs to locate the birds. A survey team typically consist of at least two people (one dog handler and one person responsible for following the transect line) and one dog. Often, more than one dog is used for a survey, but only one dog should be used at a time. The transect lines vary in length, but are typically between 1 and 8 km (range: 0.3-16.2 km, median: 3 km). When birds are observed, the exact location of observation is reported, along with its perpendicular distance from the transect line, as well as the age and sex of the birds. An observation typically includes 1 - 12 birds (mean = 5.6), with groups \> 1 typically representing one brood (female and or male with young-of-the-year chicks). When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size. Since 2019, most of the data has been collected using a mobile app tailored to the monitoring program, where the field workers can register and get access to the transect lines allocated to them by the local organizers. Prior to 2019, data were collected on a dedicated fieldwork form, and entered manually in a web portal afterwards. After field data has been registered, it undergoes several steps of quality control carried out by local stakeholders and personnel from the Norwegian Institute for Nature Research (NINA). Surveys are carried out on both public and private land. After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (<https://www.gbif.org/sampling-event-data>). The published datasets contain both metadata about the transect surveys (survey date, line transect length and location, study area ID, etc.) and bird observation data (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation). Formally, the data from public land is published as three distinct data sets, one for each of the main public land administrators (Statskog, FeFo and Fjellstyrene, respectively).

@@

The model of @nilsen2023 integrated line transect data with radio-telemetry data from from an ongoing field study of marked willow ptarmigans in Lierne municipality in Central Norway. From 2015 to 2019, around 50 birds were captured in winter (late February or early March) each year and fitted with VHF collars. The marked birds were then monitored on a regular basis until they either i) died, ii) their transmitter's battery stopped working, or iii) we lost contact with the bird for other reasons. For most of the year, the birds were monitored at least once a month by radio triangulation. Most of the fieldwork was conducted from the ground, but to avoid data gaps, the birds were also triangulated from helicopters in May, September, and November. During the breeding and chick-rearing season (May to July) birds were monitored more often, and during December and January we obtained fewer observations due to challenging field work conditions. A proportion of the birds were harvest annually in the regular recreational harvest, and birds that were harvested were reported as shot to the field personnel. In addition, as the collars had mortality switch, we were also able to locate and retrieve a high proportion of birds that died for natural causes, resulting in a known-fate mark-recapture dataset. The radio-telemetry study is described in detail in @Israelsen2020 and in @Arnekleiv2022.

@@

The three likelihoods for data resulting from the line transect sampling (observation distances, age-specific counts, and juvenile to adult ratios, see above) were also formulated as spatially explicit, with year- and area-specific distance sampling detection parameters modelled in the same way as survival and recruitment (except the effect of rodent occupancy, @fig-model). For the known-fate telemetry data (and the seasonal decomposition of survival estimated from it), on the other hand, we did not add an additional area dimension as this data was only available for on study area (the Lierne area).

@@

We implemented our multi-area IDSM in a Bayesian framework using NIMBLE version 1.0.1 [@valpine2017] in R version 4.3.1 [@R2023]. For the likelihood for line transect observation distances we used a custom half-normal distribution developed by Michael Scroggie in the "nimbleDistance" package (https://github.com/scrogster/nimbleDistance). We used non-informative uniform priors for all parameters, but used biologically sensible boundaries where possible. We simulated complete sets of initial values for all model nodes prior to model running and using pre-defined seeds to ensure reproducibility. Using NIMBLE's standard samplers, we then ran 5 MCMC chains of 150k iterations each. We discarded the first 75k samples of each chain as burn-in, and thinned the remainder by a factor 25, resulting in a final joint posterior containing a total of 5 x 3k = 15k samples (note that high thinning rates were necessary to constrain memory load of the joint posterior, which included 314568 monitored parameters).

@@

Following model fitting, we calculated posterior distributions for the proportions of variance in survival probabilities, recruitment rates, and detection decay explained by 1) spatial variation ($var_{area}$), 2) temporal variation ($var_{year}$), 3) residual variation ($var_{residual}$), and 4) variation in rodent occupancy ($var_{rodent}$). To obtain the proportion variance explained by each of the component, we divided it by the sum of all the components ($var_{area} + var_{year} + var_{residual} + var_{rodent}$). The spatial, temporal, and residual variance components were defined as the square of the estimated corresponding random effects standard deviation from the model while $var_{rodent}$ was calculated as the variance of all area- and year-specific $\beta_x*rodentOcc_{x,t}$ products. This approach for variance decomposition is equivalent to that used by @nater2018 and inspired by @nakagawa2013.

@@

All numerical results in the following are presented as median \[95% credible interval\] unless otherwise indicated. Posterior summaries (median, 95% credible interval, mean, standard deviation, coefficient of variation) for all main parameters are also provided in the supplementary file "PosteriorSummaries_byAreas.csv". Supplementary figures (SFs) are provided as .pdf files with captions in "SuppFigures_Captions.txt"; all files are deposited on OSF (<https://osf.io/7326r/>).

@@

Only during the most recent four years (2018-2021) has data been collected regularly for all reporting areas included in the analyses. During this period, estimated population densities varied between `r dens.min$post.sum` birds/km^2^ in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km^2^ in "`r dens.max$area.name`" further south near the Swedish border. In general, recent population density appeared to be lowest in the northern Norway and highest in the eastern part of central Norway (@fig-densA). Uncertainty in density estimates was relatively consistent, with a few areas (including the one with the lowest estimated density, "`r dens.min$area.name`") sticking out by having substantially less precise estimates (@fig-densB). Populations fluctuated substantially over time in any given area (SF "TimeSeries_popDens1.pdf") and some years seemed to be indicative of relatively high (e.g. 2011, 2014, 2018) or low (e.g. 2012, 2015) densities across a substantial number of areas.

@@

Annual survival probabilities ranged from `r S.min$post.sum` (area "`r S.min$area.name`") to `r S.max$post.sum` (area "`r S.max$area.name`") across reporting areas in Norway, with the highest values occurring in the far north and in the mountains in the south (@fig-pVRA). The global average survival probability across all areas and years ($\mu^S$) was estimated at `r S.hyper.mean$post.sum`. Spatial variation in survival (random effect SD = `r S.sd.spatial$post.sum`) was relatively low compared to temporal (`r S.sd.temporal$post.sum`) and residual (`r S.sd.residual$post.sum`) variation.

Recruitment rates varied between `r R.min$post.sum` (area "`r R.min$area.name`") and `r R.max$post.sum` (area "`r R.max$area.name`") and displayed a spatial pattern opposite to that of annual survival, i.e., lower recruitment rates co-occurring with higher survival rates and vice-versa (@fig-pVRB; @fig-paramCorrs B). Across all areas and years, average recruitment rate was `r R.hyper.mean$post.sum`. Unlike for survival, the model predicted similar magnitudes of spatial and temporal variation (random effect SDs of `r R.sd.spatial$post.sum` and `r R.sd.temporal$post.sum`, respectively), and about twice as much residual variation (`r R.sd.residual$post.sum`).

@@

Detection decay parameters, which determine detection probability in distance sampling surveys, varied across areas from between `r Det.min$post.sum` in "`r Det.min$area.name`" to `r Det.max$post.sum` in "`r Det.max$area.name`, resulting in detection probabilities over the transect sites ranging from `r pDet.min$post.sum` to `r pDet.max$post.sum`, respectively (truncation distance = 200 m) . The global average detection decay was `r Det.hyper.mean$post.sum` (detection probability = `r pDet.hyper.mean$post.sum`), and in general, higher values were more common in the Southern half of the country than the Northern half (SF"Avg_detect_Map.pdf"). Variation in detection over time was modest (SF "TimeSeries_pDetect.pdf"), and among-year variation in detection decay (random effect log SD = `r Det.sd.temporal$post.sum`) was almost identical in magnitude to spatial (`r Det.sd.spatial$post.sum`) and residual (`r Det.sd.residual$post.sum`) variation.

@@

Notably, the conclusive estimation of overall positive effects of rodent occupancy on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas. When @nilsen2023 fit the IDSM to data from only a single area, they were unable to obtain a reliable estimate for the rodent effect due to limited statistical power. Consequently, the space-for-time substitution that comes with extending the model across multiple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics. Doing so may also help with better understanding the mechanisms underlying the large portion of demographic rate variation that could only be attributed to random variation so far. This is the case especially for the relatively large residual variation [@fig-varDecomp] but also relevant for constant spatial and shared temporal variation. In previous work based on both marked [@Eriksen2023] and unmarked birds [@bowler2020; @Henden2020; @Novoa2016], spring conditions has come out as an important predictor of ptarmigan recruitment rates. In general, warmer and earlier springs seem to favour earlier breeding, larger clutch sizes [@Eriksen2023], and resulting higher recruitment rates measured in the late summer and early fall. @bowler2020 further reported that the strength of this relationship was not consistent in time and space, but was generally stronger in colder areas, similar to what we found for the effect of rodent occupancy here. In practice, measures representing spring conditions, such as the cover of ericaceous shrubs (a proxy for food availability) or spatially-explicit spring green up dates derived from remote-sensing data, thus constitute relevant candidate covariates for future work alongside temperature.

>>>>

Quantifying temporal and spatial variation in animal population size and demography is a central theme in ecological research and important for directing management and policy. However, this requires field sampling at large spatial extents and over long periods of time, which is not only prohibitively costly but often politically untenable. Participatory monitoring programs (also called citizen science programmes) can alleviate these constraints by recruiting stakeholders and the public to increase the spatial and temporal resolution of sampling effort and hence resulting data. While the majority of participatory monitoring programs are limited by opportunistic sampling designs, we are starting to see the emergence of structured citizen science programs that employ trained volunteers to collect data according to standardized protocols. Simultaneously, there is much ongoing development of statistical models that are increasingly more powerful and able to make more efficient use of field data. Integrated population models (IPMs), for example, are able to use multiple streams of data from different field monitoring programmes and/or multiple aspects of single datasets to estimate population sizes and key vital rates. Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale participatory monitoring program -- the "Hønsefuglportalen" -- to study spatio-temporal variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway. We constructed an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence. Recruitment rates varied more across space than over time, while the opposite was the case for survival. Slower life history patterns (higher survival, lower recruitment) appeared to be more common at higher latitudes and altitudes, portending differential effects of climate change on ptarmigan across their range. While there was variation in the magnitude of the effect small rodent occupancy had on recruitment, the relationships were predominantly positive and thus consistent with the alternative prey hypothesis. Notably, the accurate estimation of covariate effect was only made possible by integrating data from several monitoring areas for analysis. Our study highlights the potential of participatory monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns in species abundance and demographic rates, and showcases how corresponding workflows can be set up in a reproducible and semi-automated way that increases their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.

@@

While large-scale, spatially-explicit indicators for species abundance and populations are clearly needed, development and practical implementation are greatly limited due to the reliance of such indicators on the availability of data from large-scale, long-term monitoring programmes [@proencca2017]. Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last few decades. Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), the PanEuropean Common Bird Monitoring Scheme (<https://pecbms.info/>), the UK Butterfly Monitoring Scheme (<https://ukbms.org/>), the Game and Wildlife Conservation Trust Partridge Count Scheme [@aebischer2010], and the Swiss Biodiversity Monitoring program (<https://www.biodiversitymonitoring.ch/>).

@@

The line transect data from "Hønsefuglportalen" has been used previously for estimating abundance trends of willow ptarmigan (*Lagopus lagopus*) across Norway [e.g. @bowler2020; @nilsen2020], and to test a range of relevant ecological hypotheses [@bowler2020; @Breisjøberget_etal_2018]. However, large-scale estimation of demographic rates underlying abundance trends has thus far remained an untapped potential of the dataset. @nilsen2023 recently developed a novel integrated distance sampling model (IDSM) which successfully uses the age of individuals detected along line transects data coupled with radio-telemetry data to jointly estimate abundance, survival, and recruitment across years. In this study, we adapt and extend the model of @nilsen2023 to run not just on a single site but on all areas with publicly available line transect data from "Hønsefuglportalen" simultaneously. Unlike several previous studies applying integrated models for population dynamics to multiple (sub-) populations separately and comparing results [e.g. @robinson2014; @nater2023], we opt for an approach explicitly integrating across space, thus allowing for sharing of information across locations and -- in effect -- space-for-time substitution [e.g. @horswill2019; @morrison2022]. We then apply the resulting multi-area IDSM to "Hønsefuglportalen" data on willow ptarmigan to estimate population size, age-structure, survival, recruitment, and impacts of small rodent occupancy across 41 reporting districts and 15 years (2007-2021) for this culturally important small-game species. We further embed the modelling workflow in a reproducible, semi-automated pipeline that will greatly facilitate the repeated calculation of abundance and population indicators at different spatial scales as new data are added every year.

@@

The willow ptarmigan is a tetraoid bird with a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g. @fuglei2020]. While the species is currently listed as Least Concern (LC) both in the global and Norwegian Red List of Species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022]. The main reason for the long-term decline in abundance remain unresolved, but the willow ptarmigan is considered a sentinel species that is sensitive to both climate change and land use changes [@Henden2017; @Storch2007]. Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for resident predators, such as the gyrfalcon [@Alastair2020]. The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can display substantial spatio-temporal variation in demographic rates [@bowler2020]. Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycles through shared predators [@hagen1952; @bowler2020]. This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis (APH) and has been a central part of research on Fennoscandian grouse population dynamics for many decades [@Elton1942; @hagen1952; @Linden1988; @Steen1988]. In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023; @Steen1988]. Across their distributional range, willow ptarmigan are an iconic species with a high cultural value, partly linked to their popularity as game species. The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies [@Eriksen2023]. In addition, being a sentinel species, the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway, @jakobsson2020, <https://www.naturindeks.no/Indicators/lirype>] and ecosystem condition [Assessment of the Ecological Condition, @framstad2022] assessments.

@@

The line transect survey data were collected through a structured participatory monitoring program called "Hønsefuglportalen" (<https://honsefugl.nina.no/Innsyn/en>). In the first three weeks of August each year, trained volunteer fieldworkers collect observations of willow ptarmigan and other grouse species (rock ptarmigan *Lagopus muta*, black grouse *Lyrurus tetrix*, and capercaillie *Tetrao urogallus*) along predefined line transects. To increase the detection probability, fieldworkers use pointing dogs to locate the birds. A survey team typically consists of at least two people (one dog handler and one person responsible for following the transect line) and one dog. Often, more than one dog is used for a survey, but only one dog should be used at a time. The transect lines vary in length, but are typically between 1 and 8 km (range: 0.3-16.2 km, median: 3 km). When birds are observed, the exact location of observation is reported, along with its perpendicular distance from the transect line, as well as the age and sex of the birds. An observation typically includes 1 - 12 birds (mean = 5.6), with groups \> 1 typically representing one brood (female and or male with young-of-the-year chicks). When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size. Since 2019, most of the data has been collected using a mobile app tailored to the monitoring program, where the field workers can register and get access to the transect lines allocated to them by the local organizers. Prior to 2019, data were collected on a dedicated fieldwork form, and entered manually in a web portal afterwards. After field data has been registered, it undergoes several steps of quality control carried out by local stakeholders and personnel from the Norwegian Institute for Nature Research (NINA). Surveys are carried out on both public and private land. After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (<https://www.gbif.org/sampling-event-data>). The published datasets contain both metadata about the transect surveys (survey date, line transect length and location, study area ID, etc.) and bird observation data (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation). Formally, the data from public land is published as three distinct data sets, one for each of the main public land administrators (Statskog, FeFo and Fjellstyrene, respectively).

@@

The model of @nilsen2023 integrated line transect data with radio-telemetry data from an ongoing field study of marked willow ptarmigans in Lierne municipality in Central Norway. From 2015 to 2019, around 50 birds were captured in winter (late February or early March) each year and fitted with VHF collars. The marked birds were then monitored on a regular basis until they either i) died, ii) their transmitter's battery stopped working, or iii) we lost contact with the bird for other reasons. For most of the year, the birds were monitored at least once a month by radio triangulation. Most of the fieldwork was conducted from the ground, but to avoid data gaps, the birds were also triangulated from helicopters in May, September, and November. During the breeding and chick-rearing season (May to July) birds were monitored more often, and during December and January we obtained fewer observations due to challenging field work conditions. A proportion of the birds were harvest annually in the regular recreational harvest, and birds that were harvested were reported as shot to the field personnel. In addition, as the collars had mortality switches, we were also able to locate and retrieve a high proportion of birds that died from natural causes, resulting in a known-fate mark-recapture dataset. The radio-telemetry study is described in detail in @Israelsen2020 and in @Arnekleiv2022.

@@

The three likelihoods for data resulting from the line transect sampling (observation distances, age-specific counts, and juvenile to adult ratios, see above) were also formulated as spatially explicit, with year- and area-specific distance sampling detection parameters modelled in the same way as survival and recruitment (except the effect of rodent occupancy, @fig-model). For the known-fate telemetry data (and the seasonal decomposition of survival estimated from it), on the other hand, we did not add an additional area dimension as this data was only available for one study area (the Lierne area).

@@

We implemented our multi-area IDSM in a Bayesian framework using NIMBLE version 1.0.1 [@valpine2017] in R version 4.3.1 [@R2023]. For the likelihood for line transect observation distances we used a custom half-normal distribution developed by Michael Scroggie in the "nimbleDistance" package (https://github.com/scrogster/nimbleDistance, see package vignette for specifics). We used non-informative uniform priors for all parameters, but used biologically sensible boundaries where possible. We simulated complete sets of initial values for all model nodes prior to model running and using pre-defined seeds to ensure reproducibility. Using NIMBLE's standard samplers, we then ran 5 MCMC chains of 150k iterations each. We discarded the first 75k samples of each chain as burn-in, and thinned the remainder by a factor 25, resulting in a final joint posterior containing a total of 5 x 3k = 15k samples (note that high thinning rates were necessary to constrain memory load of the joint posterior, which included 314568 monitored parameters). In addition to the main model run, we implemented a second version of the model that did not use telemetry data to assess the potential impacts of auxiliary data from a single location on parameter estimates.

@@

Following model fitting, we calculated posterior distributions for the proportions of variance in survival probabilities, recruitment rates, and detection decay explained by 1) spatial variation ($var_{area}$), 2) temporal variation ($var_{year}$), 3) residual variation ($var_{residual}$), and 4) variation in rodent occupancy ($var_{rodent}$). To obtain the proportion of variance explained by each of the component, we divided it by the sum of all the components ($var_{area} + var_{year} + var_{residual} + var_{rodent}$). The spatial, temporal, and residual variance components were defined as the square of the estimated corresponding random effects standard deviation from the model while $var_{rodent}$ was calculated as the variance of all area- and year-specific $\beta_x*rodentOcc_{x,t}$ products. This approach for variance decomposition is equivalent to that used by @nater2018 and inspired by @nakagawa2013.

@@

MCMC for all central model parameters converged within the given number of iterations, but chain mixing was remained suboptimal even at convergence for some parameters. The inclusion of telemetry data from a single location into the integrated model had did not substantially affect parameter estimates beyond the seasonal decomposition of survival in Lierne, where the telemetry data was collected (see online supplementary "Comparison_noTelemetry"). All numerical results in the following are presented as median \[95% credible interval\] unless otherwise indicated. Posterior summaries (median, 95% credible interval, mean, standard deviation, coefficient of variation) for all main parameters are also provided in the supplementary file "PosteriorSummaries_byAreas.csv". Supplementary figures (SFs) are provided as .pdf files with captions in "SuppFigures_Captions.txt"; all files are deposited on OSF (<https://osf.io/7326r/>).

@@

Only during the most recent four years (2018-2021) has data been collected regularly for all reporting areas included in the analyses. During this period, estimated population densities varied between `r dens.min$post.sum` birds/km^2^ in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km^2^ in "`r dens.max$area.name`" further south near the Swedish border. In general, recent population density appeared to be lowest in northern Norway and highest in the eastern part of central Norway (@fig-densA). Uncertainty in density estimates was relatively consistent, with a few areas (including the one with the lowest estimated density, "`r dens.min$area.name`") sticking out by having substantially less precise estimates (@fig-densB). Populations fluctuated substantially over time in any given area (SF "TimeSeries_popDens1.pdf") and some years seemed to be indicative of relatively high (e.g. 2011, 2014, 2018) or low (e.g. 2012, 2015) densities across a substantial number of areas.

@@

Annual survival probabilities ranged from `r S.min$post.sum` (area "`r S.min$area.name`") to `r S.max$post.sum` (area "`r S.max$area.name`") across reporting areas in Norway, with the highest values occurring in the far north and in the mountains in the south (@fig-pVRA). The global average survival probability across all areas and years ($\mu^S$) was estimated at `r S.hyper.mean$post.sum`. Spatial variation in survival (random effect SD on logit scale = `r S.sd.spatial$post.sum`) was relatively low compared to temporal (`r S.sd.temporal$post.sum`) and residual (`r S.sd.residual$post.sum`) variation.

Recruitment rates varied between `r R.min$post.sum` (area "`r R.min$area.name`") and `r R.max$post.sum` (area "`r R.max$area.name`") and displayed a spatial pattern opposite to that of annual survival, i.e., lower recruitment rates co-occurring with higher survival rates and vice-versa (@fig-pVRB; @fig-paramCorrs B). Across all areas and years, average recruitment rate was `r R.hyper.mean$post.sum`. Unlike for survival, the model predicted similar magnitudes of spatial and temporal variation (random effect SDs on log scale of `r R.sd.spatial$post.sum` and `r R.sd.temporal$post.sum`, respectively), and about twice as much residual variation (`r R.sd.residual$post.sum`).

@@

Detection decay parameters, which determine detection probability in distance sampling surveys, varied across areas from between `r Det.min$post.sum` in "`r Det.min$area.name`" to `r Det.max$post.sum` in "`r Det.max$area.name`, resulting in detection probabilities over the transect sites ranging from `r pDet.min$post.sum` to `r pDet.max$post.sum`, respectively (truncation distance = 200 m) . The global average detection decay was `r Det.hyper.mean$post.sum` (detection probability = `r pDet.hyper.mean$post.sum`), and in general, higher values were more common in the Southern half of the country than the Northern half (SF"Avg_detect_Map.pdf"). Variation in detection over time was modest on average but the degree of temporal changes varied by area, with some areas having nearly constant detection while others showed variation by factors larger than 1.5 (SF "TimeSeries_pDetect.pdf"). The estimated average among-year variation in detection decay (random effect log SD = `r Det.sd.temporal$post.sum`) was almost identical in magnitude to spatial (`r Det.sd.spatial$post.sum`) and residual (`r Det.sd.residual$post.sum`) variation.

@@

Notably, the conclusive estimation of overall positive effects of rodent occupancy on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas. When @nilsen2023 fit the IDSM to data from only a single area, they were unable to obtain a reliable estimate for the rodent effect due to limited statistical power. Consequently, the space-for-time substitution that comes with extending the model across multiple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics. Doing so may also help with better understanding the mechanisms underlying the large portion of demographic rate variation that could only be attributed to random variation so far. This is the case especially for the relatively large residual variation ([@fig-varDecomp]) but also relevant for constant spatial and shared temporal variation. In previous work based on both marked [@Eriksen2023] and unmarked birds [@bowler2020; @Henden2020; @Novoa2016], spring conditions has come out as an important predictor of ptarmigan recruitment rates. In general, warmer and earlier springs seem to favour earlier breeding, larger clutch sizes [@Eriksen2023], and resulting higher recruitment rates measured in the late summer and early fall. @bowler2020 further reported that the strength of this relationship was not consistent in time and space, but was generally stronger in colder areas, similar to what we found for the effect of rodent occupancy here. In practice, measures representing spring conditions, such as the cover of ericaceous shrubs (a proxy for food availability) or spatially-explicit spring green up dates derived from remote-sensing data, thus constitute relevant candidate covariates for future work alongside temperature.

Another important determinant of vital rate variation is density dependence, in particular for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @aanes2002; @willebrand2001]. Negative density dependence has been found in several gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]). For willow ptarmigan, evidence for density-dependent population regulation has been mixed. @myrberget1988, for example, observed no change in productivity despite a 50% decrease in abundance, while @pedersen2004 reported strong negative density-dependence over winter and posited that dispersal may be the vital rate that responded to changes in density most strongly. Similarly, @Henden2020 reported negative density dependence when using a Gompertz-model to examine how density and a range of environmental covariates affected willow ptarmigan population dynamics in the northernmost parts of Norway. While we did not explicitly model density dependence in this study, our results can provide some preliminary insights into potential density feedbacks from both a spatial (cross-population) and a temporal (within-population) angle. Comparing average population densities and growth rate across areas did not provide evidence for strong density dependence, but there was a tendency towards the highest population growth rates appearing in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A]. When considering density dependence across years within select areas, however, we found that higher density years were associated with higher recruitment the same year, but followed by lower apparent survival probabilities and, consequently, lower population growth rates (as determined by post-hoc Pearson correlation coefficients, supplementary file "DD_corrCoef.csv"). While this seems to support the notion of negative density-dependence, testing for this post-hoc gives results that are confounded with sampling correlation [@Freckleton2006]. Hence, formally modelling density-dependence, possibly using different forms and time-lags, could prove to be a promising extension of our modelling framework in the future.
